Stressing Bank Profitability for Interest Rate Risk$^*$ \\ \vspace{0.5cm} Preliminary and Incomplete \vspace{0.5cm}

### 2013-07-17 LATEX/beg.draft.2013.tex
Revised into, edited throughout, added bibliography

\begin{abstract}
Regular bank stress testing is one of the three complementary reforms that has been made to U.S. capital regulation in response to the shortcomings of the pre-crisis regime.
The form of stress testing mandated by the Dodd-Frank Act for the Federal Reserve to undertake is one based around a small number of scenarios.  This approach involves specifying macroeconomic and financial scenarios that represents stressful conditions over the time horizon of the stress test and then projecting, for each bank holding company in the stress test, its losses, income, and path of capital \textit{given that scenario}. In this paper, we focus on the path of income. In particular, we consider the ability of empirical models to translate developments in interest rates into bank net interest margins (NIMs).
\end{abstract}

@@

\vspace{-0.1in}
Regular bank stress testing (and the capital planning that it implies) is one of the three complementary reforms that has been made to U.S. capital regulation in response to the shortcomings of the pre-crisis regime.\renewcommand{\baselinestretch}{1}\footnote{See Governor Tarullo's November 2011 speech ``The Evolution of Capital Regulation'' for this characterization of stress testing. The two other reforms that he mentions are improvements to the traditional, individual firm approach to capital regulation and the introduction of a macroprudential component to capital regulation.\vspace{0.05in}}\renewcommand{\baselinestretch}{1.5} While stress testing can take many forms---for example, being based on a small number of scenarios, being based on a large number of scenarios, or even being undertaken in reverse form to achieve a certain adverse outcome---the form of the stress tests mandated under the Dodd-Frank Act for the Federal Reserve to undertake and the form being used by the Federal Reserve for its annual Comprehensive Capital Analysis and Review (CCAR) is one based around a small number of scenarios.  This approach involves specifying a small number of macroeconomic and financial scenarios that represent stressful conditions over the time horizon of the stress test and then projecting, for each bank holding company (BHC) in the stress test, its losses, income, and path of capital \textit{given that scenario}. Accordingly, there are at least two important parts to conducting bank stress tests (i)~formulating macroeconomic and financial scenarios that should be stressful to banks and (ii)~developing econometric models that are able to translate such stressful conditions into losses, income, and capital.

We examine how well a range of time-series models can explain NIMs. We emphasize the models' pseudo-out-of-sample \textit{conditional} forecast performance. We adopt this pseudo-out-of-sample approach for the simple reason that the evaluation of models based on their in-sample forecast performance, may lead to in-sample over-fitting and poor out-of-sample performance. An explicit focus on out-of-sample performance makes forecast uncertainty explicit.
Moreover, concentrating on conditional forecasts, allows us to separate the task of modeling and forecasting NIMs from that of forecasting interest rates, which is itself a sizable topic (see Duffee, 2012).

When developing econometric models to translate stressful economic conditions into key bank financial-statement variables, it is important to consider what types of developments are likely to be stressful to banks and what particular economic variables are likely to capture these stresses. It is then critical that the models being used for stress testing be able to translate developments in these economic variables into conditional forecasts for key bank financial-statement variables.  In this paper, we examine this property for models of bank net interest margins (NIMs), defined as the ratio of net interest income (NII) to interest earning assets, where NII is the difference between a bank's interest income and interest expenses.

In determining which variables to include in our model of NIMs we look to the variables that have been emphasized by two existing strands of the literature on NIMs.  The first strand emphasizes the link between interest rates of different maturities and NIMs and is more in the macro-banking tradition. The second literature focuses on the optimal margin set by banks based on considerations like the competitive structure of the lending and deposit markets in which the bank participates, the degree of managerial risk aversion, and the degree of uncertainty associated with the future path of interest rates and is much more in the micro-banking tradition.

Our motivation for focusing on models of NIMs is fourfold. First, losses from interest rate risk have been a central feature of banking crises. Second, net interest rate margins capture income from the maturity transformation services at the center of the traditional banking model. This source of income has been lackluster given the current configuration of interest rates. Third, modelling of NIMs has received scant attention so far. Fourth, the ability of producing reliable conditional forecasts for a key component of net income is at the core of the scenario-based stress testing of banks. We elaborate on each of these four parts of our motivation below.

@@

Likewise, the Nordic banking crises of the late 1980s and early 1990s (for Finland, Sweden, and Norway)---which represent three of Reinhart's and Rogoff's~(2009) ``big five'' banking crises---were associated with elevated short-term interest rates that depressed NII and NIMs and lead to sizable bank losses and failures. In the Nordic crises, high short-term interest rates resulted from these countries attempting to maintain currency pegs with the Deutschemark in the face of the tight German monetary policy that accompanied its loose fiscal policy associated with re-unification spending. Depressed NIMs were also at the center of the U.K.'s secondary banking crisis of the early 1970s. A sudden increase in short-term interest rates abruptly drove up banks' costs of funds,  leading to extensive bank losses and failures.  To be sure, large increases in short-term rates and declines in net interest income were not the \textit{only} reason for any of these crises---indeed, there was at least one major class of asset prices that collapsed in each of these crises---but short-term interest rates---and, in particular, their elevated levels relative to long-term interest rates---were significant and in most cases equivalent in importance to the other causes of the crisis.

The recent pace of bank profit growth has been lackluster and is becoming an increasing source of concern to market commentators and regulators alike. In recent years, bank net income growth has been largely supported by unsustainable cost cutting and reductions in provisioning, rather than from sources like NII.  NII growth, for example, (in both nominal and real terms) has been well below average in recent years while NIMs (NII relative to interest earning assets) have also been depressed. Given their already low level, any adverse development that puts further downward pressure on NIMs could be very stressful for banks and, as such, it is important that the economic developments (likely involving interest rates) that would drive these outcomes are well-modeled.

Despite the importance of NIMs for the financial condition of banks and despite the attention that it receives from bank equity analysts, the modeling of NIMs has received much less attention in the risk-management literature relative to the modeling of bank variables related to credit risk.  To be sure, the recent increase in the importance of stress testing has led to the development of a number of models that link NIMs to macroeconomic variables---such as, short-term Treasury rates and the slope of the Treasury yield curve---with Kovner \textit{et al}.~(2011) and Covas \textit{et al}.~(2012) being two notable examples for the U.S.  However, this research places relatively less emphasis on the conditional forecasting properties of these models, which is important given that these models are ultimately to be used to project bank losses, income, and capital given macroeconomic scenarios.\renewcommand{\baselinestretch}{1}\footnote{Covas \textit{et al}. do present some forecast evaluation results in order to demonstrate the benefits of focusing on quantile projections. However, they only report results for net chargeoffs and pre-provisioning net revenue and their focus is on density forecast evaluation.\vspace{0.05in}}\renewcommand{\baselinestretch}{1.5}

@@

We proceed as follows. Section~2 provides a review of the literature on NIMs with an eye to cataloguing variables that have been found useful to explain the evolution of NIMs. Section~3 turns to the models that we develop to forecast NIMS. Section~4 describes the estimation data and Section~5 the estimation results. Section~6 concludes.

@@

Kovner \textit{et al}. and Covas \textit{et al.} are in the macro-banking tradition. Both link NIMs to the slope of the yield curve and a short-term interest rate.  Banks' earnings from interest income reflect two key services that banks supply---maturity transformation services (that is, borrowing short and lending long) and deposit transactions services---and the returns to these services are reflected by these two interest rate variables.  The slope of the yield curve reflects the return to banks from maturity transformation and thus is expected to enter NIM models with a positive sign. The short-term market rates in NIM models reflect the fact that bank deposit rates, while typically lower (since they provide transactions services), are constrained by the zero lower bound.  As such, the short rate places an upper limit on what banks can earn from the provision of their transactions services. As a result, the short rate is expected to enter the model, also, with a positive sign. Figure \ref{figure_nims_components} highlights the covariation of NIMs with the 3-month Treasury yield.  In the figure NIMs are split into income and expenses components, both normalized by interest earning assets. In this and in the other figures described in this section, we focus on average for the top 25 bank holding companies in the United States by assets, assessed quarterly.  The evolution of both components of NIMs bracket the evolution of the three month Treasury rate and the contours are strikingly similar. Figure \ref{figure_nims_rates} shows the evolution of NIMs against that of the 3-month and the 10-year Treasury yields.

Both Kovner \textit{et al}. and Covas \textit{et al}. measure the slope of the yield curve in their respective NIM models with the 10-year to 3-month Treasury term spread and short-term market rates with the 3-month Treasury bill rate, as do English~(2002) and English \textit{et al}.~(2012)---two other papers in this tradition.  All of these papers estimate NIM models with slightly different purposes in mind.  The NIM models estimated by Kovner \textit{et al}. and Covas \textit{et al}. are developed for stress testing and are part of much larger aggregative models of a number of components of banks' financial statements. These models are both estimated on panel data with the most notable difference between these models being the estimation techniques they employ: Least squares regression in the case of Kovner \textit{et al}. and quantile regression in the case of Covas \textit{et al}. Figure \ref{figure_nims_factors} shows the evolution of average NIMs against ``observed'' level, slope, and curvature factors.\footnote{Following Diebold and Li (2006), we compute the observed factors, respectively as: the average of the 3-month, 2-year, and 10-year Treasury yields; the difference between the 10-year and the 3-month yields; and 10-year yield minus two times the 2-year yield plus the 3 month yield.}. As made explicit by the figure, NIMs comove strongly with the slope factor over the entire sample, but the level and curvature factors also move closely with NIMs over certain subperiods.

Like our paper, English focuses exclusively on NIMs, although he does so for a number of countries (that is, all G-7 countries except for France, and for Australia, Norway, Sweden, and Switzerland).  Addditionally (and as we also do later), he considers asset yields and liability yields separately---the two measures that when differenced imply NIMs.  English's main focus is understanding how well the aggregate banking sectors in the countries he studies appear to have managed the interest rate risk that they face on their earnings, as such he uses aggregate data (rather than firm-level data) and does not focus on the conditional forecasting performance of his model.

Finally, English \textit{et al} consider the relationship between NIMs and the slope of the yield curve and short-term rates for a panel of banks, however, in contrast to the other studies they allow for the specific coefficients estimated for each bank to vary as linear functions of bank balance sheet ratios that one \textit{a priori} might think would be relevant to net interest income.  In particular, these ratios include the maturity gap between bank assets and liabilities, the deposit share of liabilities, and the loan share of assets. Again, since English \textit{et al} are focussed on understanding the determinants of NIMs they do not consider conditional forecast performance.

A separate strand of the NIM modeling literature comes from the micro-banking tradition, which focuses on the determinants of the loan rates and deposit rates that banks set as implied by the maximization of their profits. Papers in this tradition---such as, Ho and Saunders~(1981), Angbazo~(1997), and Saunders and Schumacher~(2000)---are based around what is known as the ``dealership'' model of banks; that is, banks in these models are viewed as dealers in the credit market, acting as an intermediary between demanders and suppliers of funds.  Banks are modeled as being able to raise funds from and lend funds to short-term money markets as well as being able to raise deposits from and make loans to nonfinancial agents.

In borrowing or lending in money markets banks face a market rate, which can vary over time, while in raising deposits from or extending loans to nonfinancial agents banks have the ability to set rates.  In raising deposits, banks can pay a below market rate that is fixed for the full life of the deposit and in extending loans they can earn an above market rate that is also fixed for the full maturity of the loan.  Importantly, the life of the deposit is the same as that of the loan, reflecting the fact that while core deposits can, in principal, be withdrawn on demand from a bank they are in practice extremely sticky.  Interest rate risk arises in this model from the fact that deposit-taking and lending opportunities do not arrive at the bank at precisely the same time but rather arrive at any time with some probability, where these probabilities---which are different for deposits and lending opportunities---depend on the interest rates set by the bank on deposits and loans.  Consequently, banks can find themselves holding deposits without loans, that they must then invest at variable market rates (and that could adversely fall below the deposit rate), or loans without deposits, that they must then fund at variable market rates (that could adverse rise above the lending rate).

@@

In Ho and Saunders the degree of competition faced by the bank in loan and deposit markets does not enter their estimated model, since they argue that this variable is quite slow moving and their analysis is undertaken only over a period of three years.  Saunders and Schumacher do consider the degree of competition in their analysis--which is cross-country in nature---although this takes the form of a country dummy variable in the regressions. In all cases the banks' degree of risk aversion does not enter the specification for estimated models of NIMs since this is difficult to measure, although the volatility of interest rates does and this is measured by the annual or quarterly standard deviation of weekly Treasury rates of different maturities (\textit{e.g.}, 3-month, 1-year, 2-year, 3-year, and 5-year rates). Figure \ref{figure_nims_competition} shows the evolution of NIMs against the evolution of the measure of competition that we consider -- the shadow banking sector share of assets in the overall banking sector comprising both banks and shadow banks.  For most of the sample this share increases, but the decline is arrested and partially reversed towards the end of the sample. [TO DO: add a panel to that figure that includes a measure of volatility].

Most papers in this literature also add a few more controls to the model before estimation.  These variables include banks' payments of implicit (instead of, explicit) interest on deposits;\renewcommand{\baselinestretch}{1}\footnote{This takes place through service charge remissions and other types of depositor subsidies due to regulatory restrictions of explicit interest payments and is measured by the difference between noninterest expenses and noninterest revenue, all divided by interest earning assets.\vspace{0.05in}}\renewcommand{\baselinestretch}{1.5} banks' non-interest-bearing reserve requirements;\renewcommand{\baselinestretch}{1}\footnote{The difference between noninterest expenses and noninterest revenue captures the influence of the economic cost of funds over and above the published interest expense and is measured by non-interest bearing reserves divided by total assets.\vspace{0.05in}}\renewcommand{\baselinestretch}{1.5} banks' default risk, since this should imply an additional premium charged on the rate of the banks loans given the potential loss in bank capital;\renewcommand{\baselinestretch}{1}\footnote{This is measured by net chargeoffs divided by total loans.\vspace{0.05in}}\renewcommand{\baselinestretch}{1.5} and, bank capital, since capital is costly and banks with higher capital may seek to cover this with higher lending rates.  Note that all of the empirical analysis in the ``dealership'' model literature for studying NIMs is undertaken at the firm-level and as such many of the \textit{ad hoc} controls that are added to the model are included so as to address cross-firm differences.

We would note that Angbazo does add default risk to the bank's profit-maximization problem and, as such, default risk does enter his model of NIMs in a more rigorous way.  Here he finds that for money center banks net interest margins are more sensitive to default risk than they are to interest-rate risk, while super-regional and regional banks are sensitive to interest-rate risk but less so to default risk. These findings, he notes, are consistent with the money center banks' greater concentration of short-term assets and greater use of off-balance sheet (OBS) hedging instruments.

There is a further strand of the micro-banking literature looking at determinants of banks' NIMs---which includes papers by Zarruk~(1989) and Wong~(1997)---that we mention but do not discuss at length.  In these models deposits and lending opportunities arrive at the same time and depend on the deposit and loan rates being set by banks but there is some stochastic element to the volume of deposits that banks will raise given the interest rate they set, thus requiring the banks in the model, who are risk average, to borrow or lend in the federal funds market.  These models also include features like administrative costs on deposits and loans as well as costs associated with deposit insurance premia and the primary use of these models is for comparative static exercises that examine how changes in these parameters---as well as considerations like uncertainty---will affect bank spreads.  Ultimately, however, these papers are not empirical and since the majority of their theoretical findings confirm those of those of papers like Ho and Saunders, Angbazo, and Saunders and Schumacher, we do not delve too deeply into them here.

In summing up, in addition, to short- and long-term interest rates, a few variables stand out as relevant for our analysis.  The degree of competition faced by the bank in the loan markets and the deposit markets in which it operates stands out as an important consideration, especially given that the sample we use in our analysis is much longer than the samples considered by Ho and Saunders, Angbazo, and Saunders and Schumacher.  Interest rate volatility, also appears important, since this has likely also changed over our sample.  The other variables that previous authors in the micro-banking tradition have added to their models in an \textit{ad hoc} way we hold-off for the time being from adding to our analysis.  This is largely because our analysis at this stage is at the aggregate level and these additional variables appear to have been added to explain variations in NIMs across firms.

@@

We construct forecasts from a wide array of standard methods taken from the forecasting literature. We use data with quarterly frequency and consider a forecast horizon up to ten quarters long. We are interested in forecasts for NIMs conditional on the variables in scenario $I$, which we denote $NIM_{t+s/I}$. Here are the methods considered:

\vspace{-0.1in}
\noindent \textbf{1. Equal-weighted forecast combination.} We regress NIMs on each yield $r(\tau )_{t-1}$ separately: \
\[
NIM_{t}=c_{\tau }+\rho _{\tau }NIM_{t-1}+\sum_{j=1}^4\gamma _{\tau,j}r(\tau)_{t-j}+{\eta
_{\tau ,t}}
\]
Each regression is used to produce an $s$ step-ahead forecast of NIMs conditional on yields on treasuries $r(\tau)$ in scenario $I$ with maturity $\tau$ observed through period $t+s-1.$ This conditional forecast is denoted by $NIM_{\tau,t+s/I}.$ The equal-weighted forecast combination is then given by:
\[
NIM_{t+s/I}=\sum_{\tau }\frac{NIM_{\tau ,t+s/I}}{N},
\]
where $N$ is the number of maturities considered.\footnote{Experimentation with alternative lag structures led to no improvement in the forecast results reported below.}

>>>>

\begin{abstract}
\noindent Regular bank stress testing (and the capital planning that it implies) is one of the three complementary reforms that has been made to U.S. capital regulation in response to the shortcomings of the pre-crisis regime. The form of stress testing currently undertaken by the Federal Reserve in implementing the stress testing requirement of the Dodd-Frank Act is based around a small number of scenarios.  This approach involves specifying macroeconomic and financial scenarios that represent stressful conditions over the time horizon of the stress test and then projecting, for each bank holding company in the stress test, its losses, income, and path of capital \textit{given that scenario}. In this paper, we focus on the path of income and consider the ability of empirical models to translate developments in interest rates into bank net interest margins (NIMs).
\end{abstract}

@@

\vspace{-0.1in}
Regular bank stress testing (and the capital planning that it implies) is one of the three complementary reforms that has been made to U.S. capital regulation in response to the shortcomings of the pre-crisis regime.\renewcommand{\baselinestretch}{1}\footnote{See Governor Tarullo's November 2011 speech ``The Evolution of Capital Regulation'' for this characterization of stress testing. The two other reforms that he mentions are improvements to the traditional, individual firm approach to capital regulation and the introduction of a macroprudential component to capital regulation.\vspace{0.05in}}\renewcommand{\baselinestretch}{1.5} While stress testing can take many forms---for example, being based on a small number of scenarios, being based on a large number of scenarios, or even being undertaken in reverse form to achieve a certain adverse outcome---the form of the stress tests currently being undertaken by the Federal Reserve in implementing the stress testing requirment of the Dodd-Frank Act is one based around a small number of scenarios.  This approach involves specifying a small number of macroeconomic and financial scenarios that represent stressful conditions over the time horizon of the stress test and then projecting, for each bank holding company (BHC) in the stress test, its losses, income, and path of capital \textit{given that scenario}. Accordingly, there are at least two important parts to conducting bank stress tests (i)~formulating stressful macroeconomic and financial scenarios around which stress tests can then be performed and (ii)~developing econometric models that are able to translate reliably such stressful conditions into losses, income, and capital.

We examine how well a range of time-series models can explain net interest margins---a key source of bank income generation.  In doing so, we emphasize the models' pseudo-out-of-sample \textit{conditional} forecast performance. We adopt this pseudo-out-of-sample approach for the simple reason that the evaluation of models based on their in-sample forecast performance, may lead to in-sample over-fitting and poor out-of-sample performance. Moreover, an explicit focus on out-of-sample performance makes forecast uncertainty explicit. We concentrate on conditional forecasts for a couple of reasons.  First, it allows us to separate the task of modeling and forecasting NIMs from that of forecasting interest rates, which is itself a sizable topic (see \citeN{duffee2012}).  Second, macroeconomic stress tests are in effect conditional forecasts, in this case of banking variables on stressful outcomes of aggregate macroeconomic variables.

When developing econometric models to translate stressful economic conditions into key bank financial-statement variables, it is important to consider what types of developments are likely to be stressful to banks and what particular economic variables are likely to capture these stresses. It is then critical that the models being used for stress testing are able to translate developments in these economic variables into conditional forecasts for key bank financial-statement variables.  We examine this property for models of bank net interest margins (NIMs), defined as the ratio of net interest income (NII) to interest earning assets, where NII is the difference between a bank's interest income and interest expenses.

Our motivation for focusing on models of NIMs is fourfold. First, losses from interest rate risk have been a central feature of a number of past banking crises. Second, NIMs are an important component of bank revenue generation---reflecting revenues from maturity transformation services and from deposit transactions services, two of banks' traditional activities. Given their influence on bank earnings, NIMs are a variable that receive considerable focus from bank equity analysts.\renewcommand{\baselinestretch}{1}\footnote{Indeed, one cited  rule of thumb in the industry is that 10~basis point lower NIMs means 7~percent lower earnings.  (See "Nomura Securities on U.S. Banks: Turning Japanese v2 - NIMs Hunker Down," \emph{StreetInsider.com}, June 7, 2012.) \vspace{0.05in}}\renewcommand{\baselinestretch}{1.5}   Third, the modeling of NIMs has received scant attention so far by academic researchers, despite its importance in the industry. Fourth, the ability to produce reliable conditional forecasts for a key component of net income is at the core of the scenario-based stress testing of banks. We elaborate on each of these four parts of our motivation below.

@@

Likewise, the Nordic banking crises of the late 1980s and early 1990s (for Finland, Sweden, and Norway)---which represent three of \citeN{reinhart} ``big five'' banking crises---were associated with elevated short-term interest rates that depressed NII and NIMs and lead to sizable bank losses and failures. In the Nordic crises, high short-term interest rates resulted from these countries attempting to maintain currency pegs with the Deutschemark in the face of the tight German monetary policy that accompanied its loose fiscal policy associated with re-unification spending. Depressed NIMs were also at the center of the U.K.'s secondary banking crisis of the early 1970s. A sudden increase in short-term interest rates abruptly drove up banks' costs of funds,  leading to extensive bank losses and failures.  To be sure, large increases in short-term rates and declines in net interest income were not the \textit{only} reason for any of these crises---indeed, there was at least one major class of asset prices that collapsed in each of these crises---but short-term interest rates---and, in particular, their elevated levels relative to long-term interest rates---were significant and in many cases equivalent in importance to the other causes of the crisis.

NIMs are an important component of bank revenue generation and receive considerable focus from bank equity analysts. As such, recent subdued NIM levels have become an increasing source of concern to market commentators. In recent years, bank net income growth has been largely supported by unsustainable cost cutting and reductions in provisioning, rather than from sources like NII.  NII growth, for example, (in both nominal and real terms) has been well below average while NIMs (NII relative to interest earning assets) have also been depressed. Given their already low level, any adverse development that puts further downward pressure on NIMs could be very stressful for banks and, as such, it is important that the economic developments (likely involving interest rates) that would drive these outcomes are well-modeled.

Despite the importance of NIMs for the financial condition of banks and despite the attention that it receives from bank equity analysts, the modeling of NIMs has received much less attention in the risk-management literature relative to the modeling of bank variables related to credit risk.  To be sure, the recent increase in the importance of stress testing has led to the development of a number of models that link NIMs to macroeconomic variables with \citeN{covas2012} and \citeN{hirtle2013} being two notable examples for the U.S.  However, this research places relatively less emphasis on the conditional forecasting properties of these models, which is important given that these models are ultimately to be used to project bank losses, income, and capital given macroeconomic scenarios.\renewcommand{\baselinestretch}{1}\footnote{\citeN{covas2012} do present some forecast evaluation results in order to demonstrate the benefits of focusing on quantile projections. However, they only report results for net chargeoffs and pre-provisioning net revenue and their focus is on density forecast evaluation.\vspace{0.05in}}\renewcommand{\baselinestretch}{1.5}

@@

We proceed as follows.  Section~2 provides a review of the literature on NIMs with an eye to cataloguing variables that have been found useful to explain the evolution of NIMs.  Section~3 turns to the models that we develop to forecast NIMS.  Section~4 describes the data used to estimate the forecast models and Section~5 reports our forecast results.  Section~6 uses one of the paper's models to consider the path of NIMs under the scenarios publish by the Federal Reserve as part of last year's Dodd Frank Act stress tests.  Section~7 concludes.

@@

\citeN{covas2012} and \citeN{hirtle2013} are in the macro-banking tradition and both link NIMs to the slope of the yield curve and a short-term interest rate.  Banks' earnings from interest income reflect two key services that banks supply---maturity transformation services (that is, borrowing short and lending long) and deposit transactions services---and the returns to these services are reflected by these two interest rate variables.  The slope of the yield curve reflects the return to banks from maturity transformation and thus is expected to enter NIM models with a positive sign. The short-term market rates in NIM models reflect the fact that bank deposit rates, while typically lower (since they provide transactions services), are constrained by the zero lower bound.  As such, the short rate places an upper limit on what banks can earn from the provision of their transactions services. As a result, the short rate is expected to enter the model, also, with a positive sign. Figure \ref{figure_nims_components} highlights the covariation of NIMs with the 3-month Treasury yield.  In the figure NIMs are split into income and expenses components, both normalized by interest earning assets. In this and in the other figures described in this section, we focus on average for the top 25 bank holding companies in the United States by assets, assessed quarterly.  The evolution of both components of NIMs bracket the evolution of the three month Treasury rate and the contours are strikingly similar. Figure~\ref{figure_nims_rates} shows the evolution of NIMs against that of the 3-month and the 10-year Treasury yields.

Both \citeN{covas2012} and \citeN{hirtle2013} measure the slope of the yield curve in their respective NIM models with the 10-year to 3-month Treasury term spread and short-term market rates with the 3-month Treasury bill rate, as do \citeN{english2002} and \citeN{english2012}---two other papers in this tradition.  All of these papers estimate NIM models with slightly different purposes in mind but all use essentially the same interest rate variables to summarize the yield curve.

The NIM models estimated by \citeN{covas2012} and \citeN{hirtle2013} are developed for stress testing and are part of much larger aggregative models of a number of components of banks' financial statements. These models are both estimated on panel data with the most notable difference between these models being the estimation techniques they employ: Least squares regression in the case of \citeN{hirtle2013} and quantile regression in the case of \citeN{covas2012}.

Like our paper, \citeN{english2002} focuses exclusively on NIMs, although he does so for a number of countries (that is, all G-7 countries except for France, and for Australia, Norway, Sweden, and Switzerland).  Addditionally (and as we also do later), he considers asset yields and liability yields separately---the two measures that when differenced imply NIMs.  The main focus of \citeN{english2002} is understanding how well the aggregate banking sectors in the countries he studies appear to have managed the interest rate risk that they face on their earnings, as such he uses aggregate data (rather than firm-level data) and does not focus on the conditional forecasting performance of his model.

Finally, \citeN{english2012} consider the relationship between NIMs and the slope of the yield curve and short-term rates for a panel of banks, however, in contrast to the other studies they allow for the specific coefficients estimated for each bank to vary as linear functions of bank balance sheet ratios that one \textit{a priori} might think would be relevant to net interest income.  In particular, these ratios include the maturity gap between bank assets and liabilities, the deposit share of liabilities, and the loan share of assets. Again, since \citeN{english2012} are focussed on understanding the determinants of NIMs they do not consider conditional forecast performance.

While the macro-banking NIM literature that we have just discussed relies on measures of the level and slope of the yield curve to summarize interest rates, the literature modeling the yield curve also considers the curvature.  Figure~\ref{figure_nims_factors} shows the evolution of average NIMs against a set of factors summarizing the yield curve---specifically, level, slope, and curvature factors---where these factors are all simple (additive) functions of a small number of interest rates of different maturities, similar to those used by other authors in the macro-banking literature.\renewcommand{\baselinestretch}{1.0}\footnote{The factors shown in figure~\ref{figure_nims_factors} follow \citeN{diebold2006}.  In particular, we compute the observed factors of level, slope, and curvature, respectively, as: the average of the 3-month, 2-year, and 10-year Treasury yields; the difference between the 10-year and the 3-month yields; and 10-year yield minus two times the 2-year yield plus the 3 month yield.\vspace{0.05in}}\renewcommand{\baselinestretch}{1.5} In the empirical analysis that follows we will refer to the yield curve factors that are based on simple functions of a few points on the yield  curve as ``observed'' factors.

A separate strand of the NIM modeling literature comes from the micro-banking tradition, which focuses on the determinants of the loan rates and deposit rates that banks set as implied by the maximization of their profits. Papers in this tradition---such as, \citeN{ho1981}, \citeN{angbazo1994}, and \citeN{saunders2000}---are based around what is known as the ``dealership'' model of banks; that is, banks in these models are viewed as dealers in the credit market, acting as an intermediary between demanders and suppliers of funds.  Banks are modeled as being able to raise funds from and lend funds to short-term money markets as well as being able to raise deposits from and make loans to nonfinancial agents.

In borrowing or lending in money markets banks face a market rate, which can vary over time, while in raising deposits from or extending loans to nonfinancial agents banks have the ability to set rates.  In raising deposits, banks can pay a below market rate that is fixed for the full life of the deposit and in extending loans they can earn an above market rate that is also fixed for the full maturity of the loan.  Importantly, the life of the deposit is the same as that of the loan, reflecting the fact that while core deposits can, in principal, be withdrawn on demand from a bank they are in practice extremely sticky.  Interest rate risk arises in this model from the fact that deposit-taking and lending opportunities do not arrive at the bank at precisely the same time but rather arrive at any time with some probability.  These probabilities---which are different for deposits and lending opportunities---depend on the interest rates set by the bank on deposits and loans.  Consequently, banks can find themselves holding deposits without loans, that they must then invest at variable market rates (and that could adversely fall below the deposit rate), or loans without deposits, that they must then fund at variable market rates (that could adverse rise above the lending rate).

@@

In \citeN{ho1981} the degree of competition faced by the bank in loan and deposit markets does not enter their estimated model, since they argue that this variable is quite slow moving and their analysis is undertaken only over a period of three years.  \citeN{saunders2000} do consider the degree of competition in their analysis---which is cross-country in nature---although this takes the form of a country dummy variable in the regressions. In all cases the banks' degree of risk aversion does not enter the specification for estimated models of NIMs since this is difficult to measure, although the volatility of interest rates does and this is measured by the annual or quarterly standard deviation of weekly Treasury rates of different maturities (\textit{e.g.}, 3-month, 1-year, 2-year, 3-year, and 5-year rates). Figure~\ref{figure_nims_competition} shows the evolution of NIMs against the evolution of the measure of competition that we consider -- the shadow banking sector share of assets in the overall banking sector comprising both banks and shadow banks.  For most of the sample this share increases, but the decline is arrested and partially reversed towards the end of the sample. [TO DO: add a panel to that figure that includes a measure of volatility].

Most papers in this literature also add a few more controls to the model before estimation.  These variables include banks' payments of implicit (instead of, explicit) interest on deposits;\renewcommand{\baselinestretch}{1}\footnote{This takes place through service charge remissions and other types of depositor subsidies due to regulatory restrictions of explicit interest payments and is measured by the difference between noninterest expenses and noninterest revenue, all divided by interest earning assets.\vspace{0.05in}}\renewcommand{\baselinestretch}{1.5} banks' non-interest-bearing reserve requirements;\renewcommand{\baselinestretch}{1}\footnote{The difference between noninterest expenses and noninterest revenue captures the influence of the economic cost of funds over and above the published interest expense and is measured by non-interest bearing reserves divided by total assets.\vspace{0.05in}}\renewcommand{\baselinestretch}{1.5} the rate of default risk facing the loans of the bank (since this should imply an additional premium on the rate of the banks loans given the potential loss in bank capital);\renewcommand{\baselinestretch}{1}\footnote{This is measured by net chargeoffs divided by total loans.\vspace{0.05in}}\renewcommand{\baselinestretch}{1.5} and, bank capital, since capital is costly and banks with higher capital may seek to cover this with higher lending rates.  Note that all of the empirical analysis in the ``dealership'' model literature for studying NIMs is undertaken at the firm-level and as such many of the \textit{ad hoc} controls that are added to the model are included so as to address cross-firm differences.\renewcommand{\baselinestretch}{1}\footnote{We would note that \citeN{angbazo1994} adds default risk to the bank's profit-maximization problem and, as such, default risk does enter his model of NIMs in a more rigorous way.  Here he finds that for money center banks NIMs are more sensitive to default risk than they are to interest-rate risk, while super-regional and regional banks are sensitive to interest-rate risk but less so to default risk. These findings, he notes, are consistent with the money center banks' greater concentration of short-term assets and greater use of off-balance sheet (OBS) hedging instruments.\vspace{0.05in}}\renewcommand{\baselinestretch}{1.5}

There is a further strand of the micro-banking literature looking at determinants of banks' NIMs---which includes papers by \citeN{zarruk1989} and \citeN{wong1997}---that we mention but do not discuss at length.  In these models deposits and lending opportunities arrive at the same time and depend on the deposit and loan rates being set by banks but there is some stochastic element to the volume of deposits that banks will obtain given the interest rate they set, thus requiring the banks in the model, who are risk average, to borrow or lend in the federal funds market.  These models also include features like administrative costs on deposits and loans as well as costs associated with deposit insurance premia and the primary use of these models is for comparative static exercises that examine how changes in these parameters---as well as considerations like uncertainty---will affect bank spreads.  Ultimately, however, these papers are not empirical and since the majority of their theoretical findings confirm those of those of papers like \citeN{ho1981}, \citeN{angbazo1994}, and \citeN{saunders2000}, we do not delve too deeply into them here.

In summing up, in addition, to short- and long-term interest rates, a few variables stand out as relevant for our analysis.  The degree of competition faced by the bank in the loan markets and the deposit markets in which it operates stands out as an important consideration, especially given that the sample we use in our analysis is much longer than the samples considered by \citeN{ho1981} and \citeN{saunders2000}.  Interest rate volatility, also appears important, since this has likely also changed over our sample.  The other variables that previous authors in the micro-banking tradition have added to their models in an \textit{ad hoc} way we hold-off for the time being from adding to our analysis.  This is largely because our analysis at this stage is at the aggregate level and these additional variables appear to have been added to explain variations in NIMs across firms.

@@

We construct forecasts from a wide array of standard methods taken from the forecasting literature. We use data with quarterly frequency and consider a forecast horizon up to ten quarters long. We are interested in forecasts for NIMs conditional on the variables in scenario $I$, which we denote $NIM_{t+s/I}$. This section describes the methods considered.

\noindent \textbf{1. Equal-weighted forecast combination.} We regress NIMs on each yield $r(\tau )_{t-1}$ separately: \
\[
NIM_{t}=c_{\tau }+\rho _{\tau }NIM_{t-1}+\sum_{j=1}^4\gamma _{\tau,j}r(\tau)_{t-j}+{\eta
_{\tau ,t}}
\]
Each regression is used to produce an $s$ step-ahead forecast of NIMs conditional on yields on treasuries $r(\tau)$ in scenario $I$ with maturity $\tau$ observed through period $t+s-1.$ This conditional forecast is denoted by $NIM_{\tau,t+s/I}.$ The equal-weighted forecast combination is then given by:
\[
NIM_{t+s/I}=\sum_{\tau }\frac{NIM_{\tau ,t+s/I}}{N},
\]
where $N$ is the number of maturities considered.\renewcommand{\baselinestretch}{1.0}\footnote{Experimentation with alternative lag structures led to no improvement in the forecast results reported below.\vspace{0.05in}}\renewcommand{\baselinestretch}{1.5}
